---
title: "Gene Expression Analysis: A Data Science Approach"
author: "Kishore R. Anekalla"
date: "2024-12-18"
categories: ["Data Analysis", "Bioinformatics"]
tags: ["R", "Data Visualization", "Gene Expression"]
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
    css: ../css/modern.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = "center",
  cache = TRUE
)
```

## Introduction

In this blog post, we'll explore gene expression data analysis using modern data science techniques. We'll work with simulated gene expression data to demonstrate various analytical approaches including exploratory data analysis, statistical testing, and visualization.

## Loading Required Libraries

```{r libraries}
library(ggplot2)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
```

## Generating Sample Data

For this example, we'll simulate gene expression data representing expression levels across different conditions.

```{r generate-data}
set.seed(123)
n_genes <- 100

gene_data <- data.frame(
  gene = paste0("GENE", sprintf("%04d", 1:n_genes)),
  condition_a = rnorm(n_genes, mean = 8, sd = 2),
  condition_b = rnorm(n_genes, mean = 10, sd = 2.5)
) %>%
  mutate(
    condition_a = pmax(condition_a, 0),  # Ensure non-negative
    condition_b = pmax(condition_b, 0),
    fold_change = log2(condition_b / condition_a),
    p_value = runif(n_genes, 0, 0.1),  # Simulated p-values
    log_pvalue = -log10(p_value),
    significant = p_value < 0.05
  )

head(gene_data)
```

## Exploratory Data Analysis

Let's start by examining the distribution of expression values.

```{r distribution-plot}
ggplot(gene_data, aes(x = condition_a)) +
  geom_histogram(bins = 30, fill = "#2563eb", alpha = 0.7, color = "white") +
  labs(
    title = "Distribution of Expression Values (Condition A)",
    x = "Expression Level",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )
```

## Volcano Plot

A volcano plot visualizes the relationship between fold change and statistical significance.

```{r volcano-plot}
ggplot(gene_data, aes(x = fold_change, y = log_pvalue, color = significant)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(
    values = c("FALSE" = "#6b7280", "TRUE" = "#ef4444"),
    labels = c("Not Significant", "Significant"),
    name = "Significance"
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "gray") +
  labs(
    title = "Volcano Plot: Fold Change vs. Statistical Significance",
    x = "Log2 Fold Change",
    y = "-Log10 P-value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom"
  )
```

## Heatmap of Top Differentially Expressed Genes

Let's create a heatmap showing the top 20 most differentially expressed genes.

```{r heatmap}
top_genes <- gene_data %>%
  arrange(p_value) %>%
  head(20)

heatmap_data <- top_genes %>%
  select(gene, condition_a, condition_b) %>%
  tidyr::pivot_longer(cols = c(condition_a, condition_b), 
                      names_to = "condition", 
                      values_to = "expression")

ggplot(heatmap_data, aes(x = gene, y = condition, fill = expression)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_viridis_c(name = "Expression") +
  labs(
    title = "Heatmap of Top 20 Differentially Expressed Genes",
    x = "Gene",
    y = "Condition"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 12)
  )
```

## Summary Statistics

```{r summary-stats}
summary_stats <- gene_data %>%
  summarise(
    total_genes = n(),
    significant_genes = sum(significant),
    avg_fold_change = mean(fold_change),
    high_fold_change = sum(abs(fold_change) > 1)
  )

kable(summary_stats, caption = "Summary Statistics")
```

## Key Findings

- **Total Genes Analyzed**: `r summary_stats$total_genes`
- **Significantly Differentially Expressed**: `r summary_stats$significant_genes` (p < 0.05)
- **Average Fold Change**: `r round(summary_stats$avg_fold_change, 2)`
- **Genes with |Fold Change| > 1**: `r summary_stats$high_fold_change`

## Interactive Data Table

```{r interactive-table}
DT::datatable(
  gene_data %>%
    select(gene, condition_a, condition_b, fold_change, p_value, significant) %>%
    arrange(p_value),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    order = list(list(4, 'asc'))
  ),
  caption = "Gene Expression Data",
  filter = "top"
)
```

## Conclusion

This analysis demonstrates how modern data science techniques can be applied to gene expression data. The visualizations help identify patterns that might not be obvious from raw data alone. Interactive elements allow for deeper exploration of the dataset.

## References and Further Reading

- [R Markdown Documentation](https://rmarkdown.rstudio.com/)
- [ggplot2 Documentation](https://ggplot2.tidyverse.org/)
- [Bioconductor for Genomic Data Analysis](https://www.bioconductor.org/)

